A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Revised README.md after submission

<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
  <!--
  ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)

@@

Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of independent varaibles on $y_t$. This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time. In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations. This is equivalent to specifying some effects to be invariant over time.

@@

Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24). This flexibility can be reduced and it is posible to impose restrictions to consider some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time. In this case, we can reformulate the coefficients expression ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations. This is equivalent to specifying some effects to be invariant over time.
-->

@@

<!--
![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)

>>>>

<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/ -->
  <!--
  ![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)

@@

Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of the independent variables on $y_t$. This flexibility can be reduced and it is posible to impose restrictions considering some $\beta$ coefficients as being constant over time. In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations. This is equivalent to specifying some effects to be invariant over time.

@@

Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of the independent variables on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24). This flexibility can be reduced and it is posible to impose restrictions considering some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time. In this case, we can reformulate the coefficients expression ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations. This is equivalent to specifying some effects to be invariant over time.
-->

@@

<!--
![\begin{equation} y = X \beta + \epsilon ![E\[\epsilon \epsilon'\]=\bf{\Sigma} \otimes \bf{I_N} ; \ \bf{\Sigma}=(\sigma_{tt'})](https://render.githubusercontent.com/render/math?math=E%5B%5Cepsilon%20%5Cepsilon'%5D%3D%5Cbf%7B%5CSigma%7D%20%5Cotimes%20%5Cbf%7BI_N%7D%20%3B%20%5C%20%5Cbf%7B%5CSigma%7D%3D(%5Csigma_%7Btt'%7D))

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Revised README.md after submission

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature, humidity, and
  hours of sunshine. In this research, we address this question by means
  of a spatial analysis of the incidence of COVID-19 in the provinces in
  Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach
  allows us to model the incidence of reported cases of the disease per
  100,000 population, as a function of temperature and humidity, while
  controlling for GDP per capita, population density, percentage of older
  adults in the population, and presence of mass transit systems. An
  interesting aspect of the spatial SUR approach is that it models
  incidence as a contagion process. Our results indicate that incidence of
  the disease is lower at higher temperatures and higher levels of
  humidity, although coefficients for this variable are significant only
  in some equations. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.
  \end{abstract}

@@

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
posible to impose restrinctions considering some \(\beta\) coefficients
as being constant over time. In this case, we can reformulate the
coefficients expression
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant across
equations. This is equivalent to specifying some effects to be invariant
over time.

@@

We begin our discussion with the evolution of the spatial
autocorrelation coefficient (\(\rho\)) in Figure
\ref{fig:results-time-1} (left panel). It is interesting to note that
the magnitude of the spatial autocorrelation coefficient \(\rho\)
declines over the period under analysis, and is not significant for some
days. This is an interesting result: immediately prior to the
declaration of the state of emergency, there appears to have been a
strong inter-provincial contagion effect. This situation persisted for a
few days after the lockdown, and became less strong over time. At the
end of the period under examination, the strength of this effect was
much diminished. We would expect that under full compliance with strict
lockdown conditions (meaning no inter-provincial mobility) the spatial
contagion effect would be zero. The fact that it is not is evidence of
some weak contagion effect due to restrictions to the movement between
provinces.

>>>>

  \begin{abstract}
  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
  systems and the economy. Public health interventions have slowed the
  spread of the virus, albeit at a high cost. How and when to ease
  restrictions to movement hinges in part on whether SARS-CoV2 will
  display seasonality due to variations in temperature, humidity, and
  hours of sunshine. Here, we address this by means of a spatio-temporal
  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
  Unrelated Regressions (SUR) allows us to model the incidence of reported
  cases of the disease per 100,000 population as an interregional
  contagion process, in addition to a function of temperature, humidity,
  and sunshine. In the analysis we also control for GDP per capita,
  percentage of older adults in the population, population density, and
  presence of mass transit systems. The results support the hypothesis
  that incidence of the disease is lower at higher temperatures and higher
  levels of humidity. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.
  \end{abstract}

@@

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
independent varaibles on \(y_t\). This flexibility can be reduced and it
is posible to impose restrinctions considering some \(\beta\)
coefficients as being constant over time. In this case, we can
reformulate the coefficients expression
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant across
equations. This is equivalent to specifying some effects to be invariant
over time.

@@

We begin our discussion with the evolution of the spatial
autocorrelation coefficient (\(\rho\)) in Figure
\ref{fig:results-time-1} (left panel). It is interesting to note that
the magnitude of the spatial autocorrelation coefficient \(\rho\)
declines over the period under analysis, and is not significant for some
days. This is an interesting result: immediately prior to the
declaration of the state of emergency, there appears to have been a
strong inter-provincial contagion effect. Keeping in mind that the
incubation period ranges between 2 and 11 days with a median of 5, it is
reasonable to expect that the effect of the lockdown will be observed
with some delay. Indeed, as seen in the figure, the autocorrelation
coefficient remains high for three days, then declines on March 16, and
continues to weaken over time. At the end of the period under
examination, the strength of this effect is much diminished. We would
expect that under full compliance with strict lockdown conditions
(meaning no inter-provincial mobility) the spatial contagion effect
would be zero. The fact that it is not is evidence of some weak
contagion effect due to restrictions to the movement between provinces.
